Erosion of somatic tissue identity with loss of the X-linked intellectual disability factor KDM5C

### 2024-07-22 submission/manuscript.Rmd
discussion drafting

Distinguishing between the germline (cells that pass on their genetic material, e.g. sperm and eggs) and the soma (cells that perform all other bodily functions) is a key feature of multicellularity and occurs during early embryogenesis. In mammals, chromatin regulators play a key role in decommissioning germline genes in somatic cells during the transition from naive to primed pluripotency by placing repressive histone H2A lysine 119 monoubiquitination (H2AK119ub1)[@endohPCGF6PRC1SuppressesPremature2017], histone 3 lysine 9 trimethylation (H3K9me3)[@endohPCGF6PRC1SuppressesPremature2017; @mochizukiRepressionGermlineGenes2021], and DNA CpG methylation[@velascoDnmt3bRecruitmentE2F62010; @hackettPromoterDNAMethylation2012; @mochizukiRepressionGermlineGenes2021] at germline gene promoters. Systematically characterizing KDM5C's role in germline gene repression during early embryogenesis will unveil key mechanisms underlying the demarcation between soma and germline identity and while also providing molecular footholds to test the impact of ectopic germline genes on neurodevelopment.

@@

Although KDM5C is generally thought to supress transcription through erasure of histone 3 lysine 4 di- and trimethylation (H3K4me2/3)[@iwaseXLinkedMentalRetardation2007], recent studies in ESCs have suggested KDM5C's repression _Dazl_ is independent of its catalytic activity[@guptaGenomewideScreenReveals2023]. Somatic repression of germline genes is typically established during the transition between naive and primed pluripotency, which modeled _in vitro_ as ESCs to EpiLC differentiation. In ESCs, chromatin regulators place repressive histone modifications at germline gene promoters, including histone 2A 119 monoubiquitination (H2AK119ub1) and histone 3 lysine 9 trimethylation (H3K9me3)[@endohPCGF6PRC1SuppressesPremature2017; @liuPolycombGroupProtein2020; @mochizukiRepressionGermlineGenes2021]. Germline genes are then silenced long-term in EpiLCs by _de novo_ placement of DNA CpG methylation (DNAme)[@mochizukiRepressionGermlineGenes2021]. Although KDM5C's catalytic activity is not required to suppress _Dazl_ in ESCs, KDM5C may promote long-term germline gene silencing via H3K4me3 removal since H3K4me3 can impede DNAme placement[@otaniStructuralBasisRecognition2009; @guoStructuralInsightAutoinhibition2015] and DNAme is lost at select germline gene promoters in the hippocampus[@scandagliaLossKdm5cCauses2017]. Because KDM5C's role in germline gene repression has only been characterized in ESCs and in the mature brain, it is currently unclear to what extent KDM5C is involved during transition between ESCs and EpiLCs and if its catalytic activity is required for long-term silencing.

@@

- testis genes are germline genes, germline gene list
	- While a select few tissue-enriched DEGs had known roles in the brain, the majority were testis genes with no known function outside the germline. 
	- distinguishing between germline and soma a key feature of multicellularity, chromatin regulators are important for this process
	- Studies on chromatin regulators in germline gene repression has focused on specific genes 
	- Curated a list of germline-enriched genes to robustly assess the types of genes dysregulated and KDM5C's role

@@

- We demonstrated KDM5C is important for the transition between histone-mediated to DNAme-mediated silencing of germline genes during the transition from naive to primed pluripotency.
	- Loss of DNAme can last throughout life at at least two gemrline gene promoters (hippocampus barco)
	- KDM5C could be important for DNAme is that KDM5C erases H3K4me3 which can impede CpGme.
		- In support of this, KDM5C is highly enriched at the CpG islands near germline TSS that are methylated in EpiLCs
	- However, Recently KDM5C's catalytic acivity was found to be unnecessary for dazl suppression in ESCs.  
	- Since DNAme is not placed until EpiLC stage, KDM5C's catalytic activity may be required for long-term silencing of germline genes.

>>>>

Distinguishing between the germline (cells that pass on their genetic material, e.g. sperm and eggs) and the soma (cells that perform all other bodily functions) is a key feature of multicellularity and occurs during early embryogenesis. In mammals, chromatin regulators play a key role in decommissioning germline genes in somatic cells during the transition from na\"{i}ve to primed pluripotency by placing repressive histone H2A lysine 119 monoubiquitination (H2AK119ub1)[@endohPCGF6PRC1SuppressesPremature2017], histone 3 lysine 9 trimethylation (H3K9me3)[@endohPCGF6PRC1SuppressesPremature2017; @mochizukiRepressionGermlineGenes2021], and DNA CpG methylation[@velascoDnmt3bRecruitmentE2F62010; @hackettPromoterDNAMethylation2012; @mochizukiRepressionGermlineGenes2021] at germline gene promoters. Systematically characterizing KDM5C's role in germline gene repression during early embryogenesis will unveil key mechanisms underlying the demarcation between soma and germline identity and while also providing molecular footholds to test the impact of ectopic germline genes on neurodevelopment.

@@

Although KDM5C is generally thought to supress transcription through erasure of histone 3 lysine 4 di- and trimethylation (H3K4me2/3)[@iwaseXLinkedMentalRetardation2007], recent studies in ESCs have suggested KDM5C's repression _Dazl_ is independent of its catalytic activity[@guptaGenomewideScreenReveals2023]. Somatic repression of germline genes is typically established during the transition between na\"{i}ve and primed pluripotency, which modeled _in vitro_ as ESCs to EpiLC differentiation. In ESCs, chromatin regulators place repressive histone modifications at germline gene promoters, including histone 2A 119 monoubiquitination (H2AK119ub1) and histone 3 lysine 9 trimethylation (H3K9me3)[@endohPCGF6PRC1SuppressesPremature2017; @liuPolycombGroupProtein2020; @mochizukiRepressionGermlineGenes2021]. Germline genes are then silenced long-term in EpiLCs by _de novo_ placement of DNA CpG methylation (DNAme)[@mochizukiRepressionGermlineGenes2021]. Although KDM5C's catalytic activity is not required to suppress _Dazl_ in ESCs, KDM5C may promote long-term germline gene silencing via H3K4me3 removal since H3K4me3 can impede DNAme placement[@otaniStructuralBasisRecognition2009; @guoStructuralInsightAutoinhibition2015] and DNAme is lost at select germline gene promoters in the hippocampus[@scandagliaLossKdm5cCauses2017]. Because KDM5C's role in germline gene repression has only been characterized in ESCs and in the mature brain, it is currently unclear to what extent KDM5C is involved during transition between ESCs and EpiLCs and if its catalytic activity is required for long-term silencing.

@@

A substantial portion of genes upregulated in _Kdm5c_-KO brain are testis genes that have no known function within the brain. Through the use of publically available RNA-seq datasets, we demonstrated these testis-enriched DEGs are typically unique to germ cells. Misexpression of germline genes in the brain suggests _Kdm5c_-KOs fail to demarcate between somatic and germline lineages, a key feature of multicellularity and sexual reproduction. Previous studies have demonstrated germline genes are decommissioned in somatic cells by multiple chromatin-modifying enzymes, but have primarily focused on the repression of key marker genes. To globally characterize KDM5C's role in germline gene repression, we curated a list of mouse germline-enriched genes using publically available germ cell-depleted RNA-seq datasets. This resource enabled us to identify 1) the extent of germline gene dysregulation in _Kdm5c_-KO cells, 2) the types of germline genes misexpressed at different developmental time points, and 3) which types of germline genes are directly or indirectly regulated by KDM5C.

We found misexpression of tissue-enriched genes, including germline genes, also occurs in _Kdm5c_-KO epiblast-like cells (EpiLCs), an _in vitro_ model of the post-implantation embryo.

- _In vivo_, a subset of epiblast stem cells will reactivate germline gene expression and form primordial germ cells (PGCs) that will eventually become the mature germline. This process can be mimicked _in vitro_ by differentiating EpiLCs into primordial germ cell-like cells (PGCLCs)[XXX]

- Importantly, we did not observe any significant changes in the expression of EpiLC marker genes and cellular morphology, indicating _Kdm5c_-KO EpiLCs are not simply becoming PGCLCs. 
- This is further supported proper expression of _Otx2_ in _Kdm5c_-KO EpiLCs, an epiblast stem cell marker that is known to repress differentiation into PGC and PGCLCs[@zhangOTX2RestrictsEntry2018]. This suggests aberrant germline gene expression is occuring ectopically in conjunction with typical developmental programs, in conjunction with _Kdm5c_-KO mice having overall unimpaired brain morphology and very few brain-specific DEGs.

- EpiLCs are an _in vitro_ model of the post-implantation embryo, when cells transition from na\"{i}ve to primed pluripotency.

- Only a small number of germline genes were shared between _Kdm5c_-KO EpiLCs and in the brain.

dysregulation of tissue identity

- In line with previous work on other chromatin regulators, we found dysregulation of _Kdm5c_-KO tissue identity began during early embryogenesis. While _Kdm5c_-KO epiblast-like cells (EpiLCs)

@@

- We demonstrated KDM5C is important for the transition between histone-mediated to DNAme-mediated silencing of germline genes during the transition from na\"{i}ve to primed pluripotency.
	- Loss of DNAme can last throughout life at at least two gemrline gene promoters (hippocampus barco)
	- KDM5C could be important for DNAme is that KDM5C erases H3K4me3 which can impede CpGme.

		- In support of this, KDM5C is highly enriched at the CpG islands near germline TSS that are methylated in EpiLCs
	- However, Recently KDM5C's catalytic acivity was found to be unnecessary for dazl suppression in ESCs.  
	- Since DNAme is not placed until EpiLC stage, KDM5C's catalytic activity may be required for long-term silencing of germline genes. 
	- This would be the first (?) example in which removal of an active mark is required for germline gene repression.
